A molecular taxonomy of tumors independent of tissue-of-origin

### 2021-07-02 text/introduction.Rmd
Minor edits to introduction. Add section in results regarding S5 Fig

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
genes mutated at a greater rate than expected due to chance. Dees et al. [@deesMuSiC2012]
developed a pipline to separate driver mutations from passenger mutations. 
Tamborero et al. [@tamboreroOncodriveCLUST2013] expanded on this idea to identify 
tumor mutation biases towards specific regions of a protein sequence. Lawrence et 
al. [@lawrenceDiscoverySaturationAnalysis2014] focused on identifying cancer driver 
genes with intermediate mutation frequencies using tumor-normal pairs. It has been
estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Mutations in cancer driver
genes are commonplace in healthy tissue and correlate with age and environmental
exposures [@mutations-healthy-tissues].

Other studies provide a comprehensive view of mutations, gene expression and
genomic signatures, with the goal of understanding common themes of all cancer
independent of tissue of origin. Understanding cancer as a disease of the cell
has long been a goal of the field as characterized in essays by
[@weinberg2000; @hallmarks_II]. The first of these studies in genomics identified
11 subtypes from 12 cancer types, using integrative analysis with co-equal
weighting of gene expression, methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though ~10% could be
reclassified independent of tissue-of-origin. In a second study involving 33
cancer types and a much greater number of tumors, the authors identified 28
clusters that could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous,
pan-gynecological/squamous and pan-kidney
[@hoadleyCellofOriginPatternsDominate2018].

>>>>

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
genes mutated at a greater rate than expected due to chance. Dees et al. [@deesMuSiC2012]
developed a pipeline to separate driver mutations from passenger mutations. 
Tamborero et al. [@tamboreroOncodriveCLUST2013] expanded on this idea to identify 
tumor mutation biases towards specific regions of a protein sequence. Lawrence et 
al. [@lawrenceDiscoverySaturationAnalysis2014] focused on identifying cancer driver 
genes with intermediate mutation frequencies using tumor-normal pairs. It has been
estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Mutations in cancer driver
genes are commonplace in healthy tissue and correlate with age and environmental
exposures [@mutations-healthy-tissues].

Other studies provide a comprehensive view of mutations, gene expression and
genomic signatures, with the goal of understanding common themes of all cancer
independent of tissue of origin. Understanding cancer as a disease of the cell
has long been a goal of the field as characterized in essays by Hanahan and Weinberg
[@weinberg2000; @hallmarks_II]. The first of these studies in genomics identified
11 subtypes from 12 cancer types, using integrative analysis with co-equal
weighting of gene expression, methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though ~10% could be
reclassified independent of tissue-of-origin. In a second study involving 33
cancer types and a much greater number of tumors, the authors identified 28
clusters that could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous,
pan-gynecological/squamous and pan-kidney
[@hoadleyCellofOriginPatternsDominate2018].

### 2021-07-02 text/manuscript.Rmd
Minor edits to introduction. Add section in results regarding S5 Fig

**Fig 5. Comparison of different cancers, simulated and actual results.**
Comparison of survival in six simulated cohorts. A) A group consisting of
randomly selected 30 year old women, unaffected (light blue) _vs._ cancer 
using the typical values from the posterior samples for a BRCA patient (yellow) 
and an artificial 10 years added to the effective age for a BRCA patient 
(dark blue), another group consisting of uniformly selected 30-70 year-old
women at random, unaffected (green) and cancer without (pink) and with (dark orange) 
the artificial 10 years. B) Posterior probability density of a 20 year old 
accelerated aging by cancer type (effective age). Red dashed line represents
age for unaffected individual, C) Posterior probability density of effective 
age rate for every additional year past 20 year old by cancer type. Red 
dashed line represents effective age rate for unaffected individual, D) 
Posterior probability density of the $k_{tis}$ multiplier. Values greater 
than 1 correspond to decreased age-independent life expectancy. Red dashed 
line represents $k_{tis}$ for unaffected individual. E,F,G,H) Comparison 
of our model to the actual survival for the patients in our dataset. We 
compare random group with correct age distribution for the given cancer 
(turqoise) to the model predicted survival for the given average effective 
age (blue) and the actual survival for the patients in our dataset (orange). 
Differences with actual data are likely due to a mixture of cluster 
identities which we exclude from these simulations for simplicity.

>>>>

**Fig 5. Comparison of different cancers, simulated and actual results.**
Comparison of survival in six simulated cohorts. A) A group consisting of
randomly selected 30 year old women, unaffected (light blue) _vs._ cancer 
using the typical values from the posterior samples for a BRCA patient (yellow) 
and an artificial 10 years added to the effective age for a BRCA patient 
(dark blue), another group consisting of uniformly selected 30-70 year-old
women at random, unaffected (green) and cancer without (pink) and with (dark orange) 
the artificial 10 years. B) Posterior probability density of the effective age 
baseline ($A_{0}$) by cancer type. Red dashed line represents
age for unaffected individual, C) Posterior probability density of effective 
age rate ($A_{r}$) by cancer type. Red 
dashed line represents effective age rate for unaffected individual, D) 
Posterior probability density of the $k_{tis}$ multiplier. Values greater 
than 1 correspond to decreased age-independent life expectancy. Red dashed 
line represents $k_{tis}$ for unaffected individual. E,F,G,H) Comparison 
of our model to the actual survival for the patients in our dataset. We 
compare random group with correct age distribution for the given cancer 
(turqoise) to the model predicted survival for the given average effective 
age (blue) and the actual survival for the patients in our dataset (orange). 
Differences with actual data are likely due to a mixture of cluster 
identities which we exclude from these simulations for simplicity.

### 2021-07-02 text/survival.Rmd
Minor edits to introduction. Add section in results regarding S5 Fig

>>>>

To determine the significance of our pathway-based clustering, we evaluated the 
performance in predicting time to death (tdeath) with other published classification 
models. We downloaded cluster membership form two studies conducted within the TCGA 
consortium [@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018]. 
We generated tdeath predictions from uniform random p, by inverting the cumulative 
distribution function of our death probability function. Using the actual tdeath 
from TCGA as the reference, we show that our pathway-based clustering (referred to as 
kNN) had better performance compared to COCA [@hoadleyMultiplatformAnalysis2014], 
iCluster [@hoadleyCellofOriginPatternsDominate2018] and tissue-of-origin (tis) - 
although not statistically significant (**S5A Fig**). Additionally, we computed the
tdeath error using matched sample tdeath from the prediction model and actual (**S5B Fig**). 
Again, all models had similar performance with kNN performing slightly better as 
indicated by a smaller error entropy.
